Explore Red Wine Data by Victor Lambert
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(GGally)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
wd = read.csv("wineQualityReds.csv")
# Remove the row number column, it isn't very useful
wd$X <- NULL
```

This dataset contains 1599 records of red wine chemical composition and quality
rating. The wine is the red Vinho Verde out of northern Portugal. There are 12
variables in the dataset, 11 chemical properties of the wine and quality as
reported by three different critics.

Input variables (based on physicochemical tests):
   1 - fixed acidity (tartaric acid - g / dm^3)
   2 - volatile acidity (acetic acid - g / dm^3)
   3 - citric acid (g / dm^3)
   4 - residual sugar (g / dm^3)
   5 - chlorides (sodium chloride - g / dm^3)
   6 - free sulfur dioxide (mg / dm^3)
   7 - total sulfur dioxide (mg / dm^3)
   8 - density (g / cm^3)
   9 - pH
   10 - sulphates (potassium sulphate - g / dm3)
   11 - alcohol (% by volume)
   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)

Questions we will investigate during this exploration:
- What are the variables that impact wine quality?
- Can we use the data to predict wine quality?
- Are there interesting correlations between other measurments?

# Univariate Plots Section

In this section we look into the distribution of single variables.

### Quality Distribution

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Summary}
print('Quality Summary')
summary(wd$quality)
```

It is clear from the max and min that there aren't any wines above 8 or below 3.
Also, since the first and third quantile are 5 and 6 respectively we expect the
majority of the values to fall between 5 and 6. Since it is very discrete a 
table of all the values could be useful.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Table}
print('Quality Table')
table(wd$quality)
```

Here it is clear that we should combine some bins in order to have a better
significance for those areas. We create four bins for wine quality:
- poor wines 3 and 4
- low-mediocre wines 5
- high-mediocre wines 6
- good wines 7 and 8
These bins can be used to split other variables and see if there are patterns.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_Bins}
# Cut the quality into four bins for easier use in multi-variable analysis
# Creates four bins
wd$quality.bin  <- cut(wd$quality, c(2, 4, 5, 6, 8))
print("Quality Bins Table")
table(wd$quality.bin)
```

As we expected, most of the values are either 5 or 6 for quality. There are 
quite a few 7 wines also which might be helpful for identifying trends. We will
look at the distribution in a histogram next.

```{r echo=FALSE, message=FALSE, warning=FALSE, Quality_hist}
ggplot(aes(quality), data = wd) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(2,10, 1))
```

Initially, looking at the distribution of quality we see something close to a 
normal distribution. The vast majority of the wines are in the middle of the 
spectrum, and there are not many wines at the extremes. We will try to find any
variables which can help us predict this. 

We will go through the available variables, paying particular attention to ones
which might be of particular interest.

## Acidity Metrics

```{r echo=FALSE, message=FALSE, warning=FALSE, Acidity_Plot}
p <- ggplot(data = wd)
p1 <- p + geom_histogram(aes(fixed.acidity))
p2 <- p + geom_histogram(aes(volatile.acidity))
p3 <- p + geom_histogram(aes(citric.acid))
grid.arrange(p1,p2,p3, ncol = 1)
```

Looking at the three different types of acid, the fixed and volatile acidity are
clearly close to a normal distribution. The citric acid, however doesn't look
like a normal distribution. We will look at the log of the citric acid content.

```{r echo=FALSE, message=FALSE, warning=FALSE, Citric_acid_log}
ggplot(aes(citric.acid), data = wd) +
  geom_histogram() + 
  scale_x_log10()
```

The citric acid content does seem more normal when plotted as a log, but still
not perfect since it is skewed by some low values. For the last acidity related
metric we will look at pH.

```{r echo=FALSE, message=FALSE, warning=FALSE, pH_Hist}
ggplot(aes(pH), data = wd) +
  geom_histogram()

summary(wd$pH)
```

The pH follows a very normal distribution with almost all of the values between
three and four. This likely correlates well with the acidity variables, we will
look into this more in the bi and multi variate analysis section.

### Inorganic Molecule Analysis

Next we move on to analyses of the inorganic compounds reported in the dataset.
First, we investigate the chloride content. 

```{r echo=FALSE, warning=FALSE, message=FALSE, Chloride_plot}
ggplot(aes(chlorides), data = wd) +
  geom_histogram()

summary(wd$chlorides)
```

We see that the majority of chloride concentrations lie between zero and 0.2, 
but there are some values as high as 0.6. The median value is about 0.8 mg/L. We
should zoom in on the are with most of the values to get a better idea of the 
distribution. 

```{r echo=FALSE, message=FALSE, warning=FALSE, chloride_zoom}
ggplot(aes(chlorides), data = wd) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 0.15))
```

The distribution of chlorides looks much more normal in that region. It seems to
be a normal distribution with a long tail. We will have to check if the long 
tail values influence the quality in the two variable analysis later. Now, the 
sulfur dioxide, both free and total. 

```{r echo=FALSE, message=FALSE, warning=FALSE, sulfur_plot}
freeso2 <- p + geom_histogram(aes(free.sulfur.dioxide))
totalso2 <- p + geom_histogram(aes(total.sulfur.dioxide))

grid.arrange(freeso2, totalso2)
```

The data description states that free SO2 is the type which would impact the 
flavor of the wine, particularly if it is too high. That threshold will have to 
be investigated later in the analysis. Another interesting thing to consider is 
the proportion of free over total sulfur dioxide in each wine. Perhaps that
value will be similar in all wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, sulfur_proportion}
wd$proportion.sulfur.dioxide = wd$free.sulfur.dioxide / wd$total.sulfur.dioxide
ggplot(data = wd) +
  geom_histogram(aes(proportion.sulfur.dioxide))
```

It seems like the proportion varies widely, from .02 to 0.85, with most of the 
values being between 0.25 and 0.4. Now we can print some summary statistics
for sulfur dioxide.

```{r echo=FALSE, warning=FALSE, message=FALSE, SO2_stats}
print("Free SO2")
summary(wd$free.sulfur.dioxide)
print("Total SO2")
summary(wd$total.sulfur.dioxide)
print("Proportion Free")
summary(wd$proportion.sulfur.dioxide)
```

We can see that the distributions match nicely with our observations from above.
Now we will look into the sulphate contents of the wines.

```{r echo=FALSE, message=FALSE, warning=FALSE, sulphate_plot}
ggplot(aes(sulphates), data = wd) + 
  geom_histogram()
```

From this histogram it seems that the sulphates follow a similar long tail 
distribution to many of the other salts contained in the wine. We will have to
check later if the higher values have a larger impact on quality.

### Bulk Property Analysis

Now, we will move on to the analysis of some bulk properties of the wine 
including sugar content, alcohol content, and density. First, we look at the 
distribution of residual sugar content. This is the amount of sugar remaining
after the fermentation of the wine. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Sugar_Plot}
p5 <- p + geom_histogram(aes(residual.sugar))
p5
```

When looking at the sugar data, we see a simlar pattern to many other wines. 
Most fall within a lower bell curve and there is a long tail of much sweeter
wines. The sugar content of the final wine is usually set by the winemaker, so 
it is possible that most people are targeting a dryer wine in that particular
RS region. We can split the data across different quality bins to see if we can
find a pattern.

```{r echo=FALSE, message=FALSE, warning=FALSE, sugar_split}
p5 + facet_wrap(~quality.bin, scales = 'free') +
  scale_x_continuous(limits = c(0, 16))
```

There is not apparent a trend with sugar content and quality in the different
groups of wines. Next we look at the distribution of alcohol contents.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Plot}
ggplot(aes(alcohol), data = wd) +
  geom_histogram(bins = 60)
```

Unlike the sugar content and many other variables, the alcohol content does not
follow the long tail distribution. It seems that the alcohol distribution has a 
many more values in the long tail. This could be a result of the differing 
ripeness of grapes at harvest. If they are targeting a particular residual sugar
value, the alcohol content will be primarily constrained by the initial sugar
content of the grapes. We can split the alcohol content along different
quality groupings as well.

```{r echo=FALSE, message=FALSE, warning=FALSE, Alcohol_Split}
ggplot(aes(alcohol), data = wd) +
  geom_histogram() +
  facet_wrap(~quality.bin, scales = 'free') +
  scale_x_continuous(limits = c(8, 15))
```

Split over the four quality bins, it is clear that higher alcohol seems to
correlate with better quality. We will have to investigate this relationship
further. The alcohol is not necessarily the cause of the improvement, it could
be that riper grapes or some other variable improve quality and higher alcohol
is a byproduct of those. The final variable for us to investigate is density.

```{r echo=FALSE, message=FALSE, warning=FALSE, Density_Plot}
ggplot(aes(density), data = wd) +
  geom_histogram()
```

The density follows a very normal distribution, it should be closely related 
with alcohol and sugar content.

# Univariate Analysis

We are able to draw some useful information from looking at the single variable
analysis. We can see that the alcohol content likely correlates with better 
quality in the wine.

### What is the structure of your dataset?

The data contains 12 variables, 11 of which are floating point numbers and one
is an integer. It also contains an id, called 'X' which has been converted to 
a factor to prevent numerical analysis. The data already is tidy and does not
have missing values or many other issues.

### What is/are the main feature(s) of interest in your dataset?

The dataset consists of one important metric, the quality of the wine, and 11
other chemical variables about the wine. The main feature of interest is how
we can predict the quality of the wine from some combination of these other
chemicals.

### What other features in the dataset do you think will help support your 
investigation into your feature(s) of interest?

The dataset contains many interrelated variables, we will also look to see how
these variables vary together to supplement the quality based analysis.

### Did you create any new variables from existing variables in the dataset?

Yes, two new variables were created. The first, quality.bin groups the 3 and 4 
values together, and the 7 and 8 values together. This is to increase the
number of wines in the high quality and low quality bins for the purpose of 
data splitting. The second created variable is the proportion of SO2 that is 
free, which didn't prove to be particularly helpful yet.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

There were many variables which had long tail distributions, although that was
the most unusual thing found so far. No further tidying or adjustment has been
done on the data. Now, we will move on to analysis of two variables.

# Bivariate Plots Section

First, we can take a look at a ggpairs plot to see if there are any unexpected
relationships that we should investigate later. The plot shows correlations
on the upper half, frequency polygons on the diagonal, and scatter plots on the 
lower diagonal.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=10, ggpairs_plot}

ggpairs(subset(wd, select = -c(quality.bin, proportion.sulfur.dioxide)),
        lower = list(continuous = 'smooth'))
```

Looking at the ggpairs, we can see some expected relationships and some 
unexpected ones. Several variables have correlations with quality, although none
of the correlations are large. The biggest effect is alcohol with a correlation
of 0.48. The other larger correlations are volatile acidity (-0.39), 
sulphates (0.25) and citric acid (0.22). This makes sense from what we know 
about these acids, citric acidhas a pleasant flavor but acetic (volatile) acid 
does not.

As far as other variables go, we can see that the different acid variables and 
pH are also correlated, although the volatile acididy correlation is the 
opposite of what we would expect since lower pH is more acidic. It also seems 
that acidict has a relationship with density which I did not predict in the
univariate analysis.

Let's take a closer look at how some of these things relate to quality, starting
with alcohol. Note that when plotting with quality as the dependent variable
adding jitter and some transparancy is very helpful to visualize the data.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol_scatter}
ggplot(aes(x = alcohol, y = quality), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')
```

Here we an see the relationship that was hinted at in the univariate analysis.
Wines with higher alcohol definitely have a higher quality average. The effect
is not strong, but there are not many better than average wines for us to draw
conclusions from.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol_box}
ggplot(aes(x = quality.bin, y = alcohol), data = wd) +
  geom_boxplot(alpha = 1/5)
```

Taking a look at this from a different plot, we can see that the higher quality
wines definitely have a higher value for alcohol. The box plot is nice because
it shows us the median as well as some grouping statistics.

```{r echo=FALSE, message=FALSE, warning=FALSE, citric_scatter}
ggplot(aes(x = citric.acid, y = quality), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(method = 'lm', color = 'black')
```

Looking at the citric acid, we can see that the correlation is small. It seems
that there are more high quality wines with higher citric acid, and more low
quality wines with low citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE, volatile_scatter}
ggplot(aes(x = volatile.acidity, y = quality), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')
```

For volatile acid, the relatinship is more clear. Having more volatile acid is
detrimental to wine quality. Now let's take a look at the sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE, sulphiate_scatter}
ggplot(aes(x = sulphates, y = quality), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')
```

I wasn't expecting the sulphates to have a positive relationship. Sulphates can
have a preservative effect, so perhaps they are preserving some of the fresh 
flavor of the wines. Let's take a look at the sulfur dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE, sulfurdioxide_scatter}
p6 <- ggplot(aes(x = free.sulfur.dioxide, y = quality), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')

p7 <- ggplot(aes(x = total.sulfur.dioxide, y = quality), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')

grid.arrange(p6, p7, ncol = 2)
```

This result is somewhat unexpected. We do expect sulfur dioxide to impact 
quality, since higher concentrations can give an off smell. However, the data
documentation says that only free sulfur dioxide should impact the smell.
However, it seems like total sulfur dioxide has a much bigger impact on the 
quality rating of the wine. Let's move on toconfirm some relationships that we 
don't expect to have an effect.

```{r echo=FALSE, message=FALSE, warning=FALSE, sugar_scatter}
no_sugar_outliers = subset(wd, residual.sugar < 4)
ggplot(aes(x = residual.sugar, y = quality), data = no_sugar_outliers) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')
cor.test(wd$residual.sugar, wd$quality)
```

As we expected from the univariate analysis, sugar content doesn't seem to have
any impact at all on the quality of the wine. The calculated correlation is low,
0.014. Now we can take a look at relationships between some other variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, density_sugar}
ggplot(aes(x = residual.sugar, y = density), data = wd) +
  geom_jitter(alpha = 1/5)
```

We can see that the density has a positive correlation with sugar, as we would 
expect. It seems that there are many outliers, so let's take a look at the area
of the plot with most of the data.

```{r echo=FALSE, message=FALSE, warning=FALSE, density_sugar_zoom}

ggplot(aes(x = residual.sugar, y = density), 
       data = no_sugar_outliers) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')
```

We can see that the trend is similar, however the sugar content doesn't explain
all the variation in density. For future analysis involving sugar we will
eliminate the outliers in a similar procedure.

```{r echo=FALSE, message=FALSE, warning=FALSE, density_alcohol}
ggplot(aes(x = alcohol, y = density), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')
```

Taking a look at the effect that alcohol has on density, it's clear that the 
lower density of ethanol is decreasing the density of lots of the wines. It's
likely that alcohol and sugar explain much of the density variation for the 
wines, we an look at them all together in the multivariate analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE, pH_fixed_scatter}
p8 <- ggplot(aes(x = pH, y = fixed.acidity), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')

p9 <- ggplot(aes(x = pH, y = volatile.acidity), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')

grid.arrange(p8, p9, ncol = 2)
```

Taking a look at the acidity, we see that fixed acidity has a much higher effect
on pH than volatile acidity. We actually see the opposite effect in volatile
acidity, which is unexpected. Possibly the fixed acidity, which has a much 
larger magnitude is affecting the pH more.

```{r echo=FALSE, message=FALSE, warning=FALSE, volatile_fixed_scatter}
ggplot(aes(x = fixed.acidity, y = volatile.acidity), data = wd) +
  geom_jitter(alpha = 1/5) +
  geom_smooth(color = 'black', method = 'lm')
```

Taking a look at fixed acidity vs volatile acidity, we do see that wines with
more volatile acidity tend to fall lower on the fixed acidity scale. This
probably explains a large portion of the unexpected pH trend. Another thing 
we can take a look at is the relationship between sulphates and sulfur dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE, so2_sulphite}
p11 <- ggplot(aes(x = sulphates, y = free.sulfur.dioxide), data = wd) +
  geom_point(alpha = 0.2)

p12 <- ggplot(aes(x = sulphates, y = total.sulfur.dioxide), data = wd) +
  geom_point(alpha = 0.2)

grid.arrange(p11, p12, ncol = 2)

cor.test(wd$total.sulfur.dioxide, wd$sulphates)
cor.test(wd$free.sulfur.dioxide, wd$sulphates)
```

Though the data description suggests that sulphates can have a large impact
on the amount of sulfur dioxide, the data suggests there is not much 
correlation.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

There are several factors which impact the quality rating of the wine. Higher
alcohol content was once again shown to be correlated with a higher quality of
the wine. Some other variables were also associated with better quality, 
including citric acid and sulphate content. The sulphate content was surprising 
to me, but could be a result of it acting as a preservative.

We also saw that volatile acidity and total sulfur dioxide content had a 
negative correlation with quality rating in wine. The effect of total sulfur
dioxide was bigger than the effect of free sulfur dioxide, which was unexpected.

Overall, alcohol is the strongest positive predictor of quality, however there
are not any very strong correlations for the quality rating.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I also looked into the relationship between some bulk properties of the wine,
particularly the density. It seems like the alcohol content is a large predictor
of lower density, which follows according to the lower density of ethanol. As a 
side note, as a chemical engineer I know that this relationship is not linear,
though at these concentrations it shouldn't be a big deal. The sugar content 
also has an impact on the density, though not as large as the alcohol content.

pH and the contribution of fixed and volatile acidity were also examined. There
was one surprising result of volatile acidity, but it seems to be explained
by the higher magnitude of the fixed acidity.

### What was the strongest relationship you found?

The strongest correlation I found is the relationship between acidity and pH 
with a correlation of 0.68. Density and alcohol also have a very high 
correlation. The strongest correlation for the primary feature of the data
is the alcohol content, with a correlation of 0.48.

# Multivariate Plots Section

One thing I want to investigate here is the idea of balance in the flavor of 
wine. Balance is the amount of acidity which counters the sweetness of the wine,
and it should be possible to look at the ratio of different types of acid and 
residual sugar to see if there is an impact on the perceived quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, balance_1}
p9 <- ggplot(aes(x = fixed.acidity, 
           y = residual.sugar, 
           color = quality.bin), data = wd) +
  geom_point(alpha = 0.5) +
  scale_color_brewer() +
  theme_dark()

p10 <- ggplot(aes(x = volatile.acidity, 
           y = residual.sugar, 
           color = quality.bin), data = wd) +
  geom_point(alpha = 0.5) +
  scale_color_brewer() +
  theme_dark()

grid.arrange(p9, p10)
```

Plotting the sugar and different types of acidity on a scatter plot with 
different colors for quality bins, it doesn't seem clear initially that there
is a relationship. Let's take a look at some box plots.

```{r echo=FALSE, message=FALSE, warning=FALSE, balance_2}
ggplot(aes(x = quality.bin, 
           y = residual.sugar/fixed.acidity),
       data = no_sugar_outliers) +
  geom_boxplot()
```

Looking at the box plots of fixed acidity vs quality, we see that there isn't
much of a relationship. This means that the balance of these wines probably
isn't very different for each quality. Let's take a look at the volatile
acidity next.

```{r echo=FALSE, message=FALSE, warning=FALSE, balance_3}
ggplot(aes(x = quality.bin, 
           y = residual.sugar/volatile.acidity),
       data = no_sugar_outliers) +
  geom_boxplot()
```

Looking at the sugar/volatile acidity, there definitely seems to be a 
relationship here. However, we know already that wines with lower volatile
acidity tend to have higher quality. This could be the cause of the relationship
we see here. We can calculate the correlation of these different variables with
quality as a test.

```{r echo=FALSE, message=FALSE, warning=FALSE, balance_stats}
cor.test(wd$volatile.acidity, wd$quality)
cor.test(wd$residual.sugar/wd$volatile.acidity, wd$quality)
```

Here we see that the lower volatile acidity is more strongly correlated with
higher quality than increased sugar/volatile acid ratio. We can't say for sure
which variable causes the effect but it is possible that the balance is not
the primary cause and we are only seeing the impact of high volatile acidiy. In
summary, it seems like we can't a significant correlation between the 'balance'
of the wine and its quality rating.

Next we will move on to analyzing the quality of the wine with respect to
alcohol content, trying to see if any variables have a synnergystic effect.
First, we look at citric acid and alcohol with the points colored by quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol_multi_1}
ggplot(aes(x = alcohol, y = citric.acid, color=quality.bin), data = wd) +
  geom_point(alpha = 1/2) +
  scale_color_brewer() +
  theme_dark()
```

Looking closely at this plot, we can see that many of the high quality wines
seem to have both higher alcohol and higher citric acid. Also, many of the lower
quality wines cluster in the bottom left, so the lower of both quantities also
isn't beneficiial for wine quality.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol_multi_2}
ggplot(aes(x = 0.75 * round(alcohol / 0.75), 
           y = citric.acid, 
           color=quality.bin), 
       data = wd) +
  geom_jitter(alpha = 1/2) +
  scale_x_continuous(limits = c(8.5, 14)) +
  geom_line(stat = 'summary', fun.y = mean) +
  scale_color_brewer() +
  theme_dark()
```

Taking a look at a slightly more complex version of the plot, we can see the
variation for different bins of wine quality. It definitely seems like the
higher citric acid is beneficial, as well as the higher alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol_multi_3}
ggplot(aes(x = alcohol, y = sulphates, color=quality.bin), data = wd) +
  geom_point(alpha = 1/2) +
  scale_y_continuous(limits = c(NA, 1.25)) +
  scale_x_continuous(limits = c(8.5, 14)) +
  geom_smooth(se = FALSE) +
  scale_color_brewer() +
  theme_dark()
```

We see a similar relationship when we look at sulphates, alcohol, and quality.
This is not surprising since alcohol and sulphates are not correlated, but they
do correlate with higher quality ratings.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol_multi_4}
ggplot(aes(x = alcohol, 
           y = total.sulfur.dioxide, 
           color=quality.bin), 
       data = wd) +
  geom_point(alpha = 1/2) +
  scale_color_brewer() +
  theme_dark()
```

Taking a look at the total sulfur dioxide, alcohol, and quality. We expect to
see a threshold above which the wine quality starts to suffer, however this does
not seem to be the case. Alcohol content is a much stronger predictor than 
sulfur dioxide content.

```{r echo=FALSE, message=FALSE, warning=FALSE, alcohol_multi_5}
ggplot(aes(x = alcohol, 
           y = free.sulfur.dioxide, 
           color=quality.bin), 
       data = wd) +
  geom_point(alpha = 1/2)+
  scale_color_brewer() +
  theme_dark()
```

Though the data documentation suggests that wines over 50ppm free sulfur dioxide
will siffer, that doesn't seem to be the case here (although there aren't many
samples with SO2 free > 50ppm). 

Next, we will take a look at the density relationship again to see if we can
explain it further. We had previously seen that there was a close relationship
with alcohol content and density. We can add the acidity into that relationship.

```{r echo=FALSE, message=FALSE, warning=FALSE, denisty_multi}
ggplot(aes(x = alcohol, 
           y = density, 
           color=fixed.acidity), 
       data = wd) +
  geom_jitter(alpha = 0.5) +
  scale_color_gradient(low = 'white', high = 'black')
```

Looking at this plot, the fixed acidity explains a large part of the additional
variation in density!

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Most of the individual relationships related to the quality of the wine
are fairly weak. Even when looking in to relationships of acidity, sugar,
and many other variables paired with alcohol no significantly stronger 
correlations are found.

When combining different variables you can see regions of the graph for 
different quality bins, however the effects do not seem to increase the 
strength of the other.

### Were there any interesting or surprising interactions between features?

One relationship that was particularly strong relative to the others is the
relationship between density, alcohol, and fixed acid. This makes scientific 
sense because ethanol is less dense than water, but some acids are more dense.
Looking at them in a group of three, there is clear trend both with alcohol 
decreasing the density but fixed acidity increasing the density.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(quality), data = wd) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(2,10, 1)) +
  labs(x = "Quality",
       y = "Count",
       title = "Quality Distribution")
```


### Description One

The first very important plot, although it is simple, is the distribution of 
quality score (1-10) in the dataset. We can see that there are very few wines 
with quality scores of 3, 4, 7, and 8. This shows that the wines follow a 
somewhat normal distribution in quality, however we have very little signal on 
what makes an excellent or extremely poor wine. This is important because it
limits the future analysis that is possible, there is significant noise on most
factors and without many points in the high or low rating it is very difficult
to find strong correlations.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = alcohol,
           y = quality,
           color = quality.bin),
       data = wd) +
  geom_jitter(alpha = 0.5) +
  labs(x = 'Alcohol [%]',
       y = 'Quality',
       color = 'Quality Bin',
       title = 'Alcohol vs. Quality') +
  theme_dark() +
  scale_color_brewer(palette = "Reds")
```

### Description Two

There is one relationship related to the quality of wine that was moderately
strong, higher alcohol wines tend to get higher quality scores. Looking at
this plot it is clear that higher scoring wines also have higher percentages of
alcohol. Lower scoring wines tend to be between 9 and 12 percent, while high
scoring wines reach as high as 14 percent. Although this is the strongest
relationship in the dataset, it is only a moderate correlation with a pearson R
of about 0.47.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = alcohol, 
           y = density, 
           color=fixed.acidity), 
       data = wd) +
  geom_jitter(alpha = 0.5) +
  scale_color_gradient(high = 'dark blue', low = 'grey') +
  labs(x = 'Alcohol [%]',
     y = 'Density [g*cm^-3]',
     color = 'Fixed Acidity [g / L]',
     title = "Density vs. Alcohol and Acidity")
```

### Description Three

The last plot shows something not related to quality, but how concentrations of
fixed acids and alcohol impact the density. These are some of the strongest
relationships in the dataset, showing how increasing alcohol decreases the 
density of the wine but increasing acidity increases the density. These
correlations are clear from the scatter plot where the gradient color of the
point corresponds to acidity.

------

# Reflection

As a wine lover and data analyst, I was really hoping to find some very 
clear and interesting relationships in this dataset. However, the low number
of high and low rated wines made it difficult to identify strong effects from
a variety of factors. There are several things which do impact the wine quality,
but only slightly. 

With this in mind, the dataset is not extremely useful from the perspective of
a chemical engineer trying to find compounds to increase or tweaks to wine 
contents (other than increase the alcohol). It seems like it could be more
suited for creating a model that can try to predict the quality of wine from
a few chemical inputs. However, the relationships are weak so the model would
benefit from significanly more samples. Perhaps the white wine database, which
has about 5000 samples, could be used to do a stronger analysis.

There were several surprising things in this dataset, the first of which was the
general lack of strong correlations. This made it difficult for me to direct
my exploration and left me often just checking lots of variables and not finding
a strong trend. However, the data was numeric for lots of variables which 
allows for creation of many nice box and scatter plots. Overall, the exploration
went well but could easily be improved in the future.