Explore Red Wine Data by Victor Lambert
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wd = read.csv("wineQualityReds.csv")
# Update the X column to a factor
wd$X <- factor(wd$X)
```

> **Tip**: Before you create any plots, it is a good idea to provide a short
introduction into the dataset that you are planning to explore. Replace this
quoted text with that general information!

This dataset contains 1599 records of red wine chemical composition and quality
rating. 

Input variables (based on physicochemical tests):
   1 - fixed acidity (tartaric acid - g / dm^3)
   2 - volatile acidity (acetic acid - g / dm^3)
   3 - citric acid (g / dm^3)
   4 - residual sugar (g / dm^3)
   5 - chlorides (sodium chloride - g / dm^3)
   6 - free sulfur dioxide (mg / dm^3)
   7 - total sulfur dioxide (mg / dm^3)
   8 - density (g / cm^3)
   9 - pH
   10 - sulphates (potassium sulphate - g / dm3)
   11 - alcohol (% by volume)
   Output variable (based on sensory data): 
   12 - quality (score between 0 and 10)

Questions we will investigate during this exploration:
- What are the variables that impact wine quality?
- Can we use the data to predict wine quality?
- Are there interesting correlations between other measurments?

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

## Quality Distribution

```{r echo=FALSE, Univariate_Plots}
print('Quality Summary')
summary(wd$quality)
```

It is clear from the max and min that there aren't any wines above 8 or below 3.
Also, since the first and third quantile are 5 and 6 respectively we expect the
majority of the values to fall between 5 and 6. Since it is very discrete a 
table of all the values could be useful.

```{r echo=FALSE, Univariate_Plots}
print('Quality Table')
table(wd$quality)
```

As we expected, most of the values are either 5 or 6 for quality. There are 
quite a few 7 wines also which might be helpful for identifying trends. We will
look at the distribution in a histogram next.

```{r echo=FALSE, Univariate_Plots}
summary(wd$quality)
ggplot(aes(quality), data = wd) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(2,10, 1))
table(wd$quality)
```

Initially, looking at the distribution of quality we see something close to a 
normal distribution. The vast majority of the wines are in the middle of the 
spectrum, and there are not many wines at the extremes. We will try to find any
variables which can help us predict this. 

We will go through the available variables, paying particular attention to ones
which might be of particular interest.

## Acidity Metrics

```{r echo=FALSE, Univariate_Plots}
p <- ggplot(data = wd)
p1 <- p + geom_histogram(aes(fixed.acidity))
p2 <- p + geom_histogram(aes(volatile.acidity))
p3 <- p + geom_histogram(aes(citric.acid))
grid.arrange(p1,p2,p3, ncol = 1)
```

Looking at the three different types of acid, the fixed and volatile acidity are
clearly close to a normal distribution. The citric acid, however doesn't look
like a normal distribution. We will look at the log of the citric acid content.

```{r echo=FALSE, Univariate_Plots}
ggplot(aes(citric.acid), data = wd) +
  geom_histogram() + 
  scale_x_log10()
```

The citric acid content does seem more normal when plotted as a log, but still
not perfect since it is skewed by some low values. For the last acidity related
metric we will look at pH.

```{r}
ggplot(aes(pH), data = wd) +
  geom_histogram()
```

The pH follows a very normal distribution with almost all of the values between
three and four. This likely correlates well with the acidity variables, we will
look into this more in the bi and multi variate analysis section.

## Inorganic Molecule Analysis

```{r}
ggplot(aes(chlorides), data = wd) +
  geom_histogram()
```

```{r}
freeso2 <- p + geom_histogram(aes(free.sulfur.dioxide))
totalso2 <- p + geom_histogram(aes(total.sulfur.dioxide))

grid.arrange(freeso2, totalso2)
```


```{r}
ggplot(aes(sulphates), data = wd) + 
  geom_histogram()
```


## Bulk Property Analysis
```{r echo=FALSE, Univariate_Plots}
p5 <- p + geom_histogram(aes(residual.sugar))
p5
```

```{r echo=FALSE, Univariate_Plots}
p5 + scale_x_continuous(limits = c(0,4))
summary(wd$residual.sugar)
```

```{r}
ggplot(aes(alcohol), data = wd) +
  geom_histogram()
```

```{r}
ggplot(aes(density), data = wd) +
  geom_histogram()
```


# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!